Exploring the Impact of Atypical Events on Regret: Analysis Through Hitchhiker, Car Accident, and Robbery Scenarios

### 2024-02-14 outputs/paper/paper.qmd
new update

---
title: "Exploring the Impact of Atypical Events on Regret: Analysis Through Hitchhiker, Car Accident, and Robbery Scenarios"
subtitle: "Reproduction of 'The impact of past behaviour normality on regret: replication and extension of three experiments of the exceptionality effect' (Lucas Kutscher & Gilad Feldman, 2018)"
author: Dailin(Ben) li
thanks: "Code and data are available at: https://github.com/UofT-DailinLi/Impact-of-Atypical-Events-on-Regret.git. A replication of various aspects in this paper are available at: https://doi.org/10.1080/02699931.2018.1504747"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This study aims to replicate Kutscher and Feldman’s findings, which examine regret based on the norm theory. According to norm theory, an individual’s activity can be categorized as either typical or atypical， with the latter, on the other hand, eliciting a stronger emotional response. We focused on replicating three specific scenarios: hitchhiking, car accidents, and robberies. Examining how people recalled their behaviour in these specific scenarios helped our perception of the emotion of regret. This replication is consistent with the original study."

@@

toc: true
toc_depth: 2
number-sections: true
bibliography: references.bib
---
\newpage
```{r echo=FALSE, message=FALSE}

@@

<!-- Background / Broader Context -->
Regret is an uncomfortable state of mind that encompasses feelings of sorrow and distress over unfortunate events, perceived personal failures, losses, violations, deficiencies, or errors[@gilovich1995experience]. It associates to many things in our daily life. Matters of regret can include both acts done and opportunities missed; they may extend from intentional actions to those that are involuntary or happenstance; they could be tangible actions carried out or purely thoughts, whether self-initiated or by others; and they might constitute ethical or legal violations, or be actions that are neutral in terms of morality and law [@gilovich1995experience].

@@

* Hitchhiker-scenario
* Car accident scenario
* Robbery scenario

@@

All of the data for REGRET and COMPENSATION contained 344 observations, and 21 VARIABLES were taken from the data underlying Kutscher and Feldman's study[@kutscher2019impact], which was collected as a questionnaire. The questionnaire recipients read a scenario in which a robbery occurred at a grocery store. In the first scenario (routine behaviour), Mr. Paul visited the store he frequented. In the second scenario (self-produced exception), the main character went to another store because he wanted a change of scenery. In the third scenario (other produced exception), Mr. Paul is forced to patronize another store because the store he frequents is closed for construction, and he runs into a robbery taking place in the store. In the next question, we ran a compensation measure followed by a regret measure. Respondents were asked how much money Mr. Paul should receive in compensation for his loss—scale from 0 to 10: 0$ to 1000000$. The next question was: assume no compensation was given to Mr. Paul. How much regret does he feel over the situation? Scale 1 from no regret to 5, very strong regret.

## Attributes

@@

## fig1
```{r echo=FALSE, message=FALSE, fig.cap="Proportions for perceived regret, injunctive social norms, descriptive social norms, and negative affect", label= "fig-1"}
fig1_data <- data.frame(
  Scenario = c("Smith - Routine", "Jones - Exception"),
  PerceivedRegret = c(7.9, 92.1),
  InjunctiveSocialNorms = c(4.7, 95.3),
  DescriptiveSocialNorms = c(90.6, 9.4),
  NegativeAffect = c(7.3, 92.7)
)

@@

p <- ggplot(data_long, aes(x = variable, y = value, fill = Scenario)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.7), width = 0.7) +
  scale_y_continuous(labels = scales::percent) +

  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'top')

@@

## fig3

@@

>>>>

---
title: "Exploring the Impact of Atypical Events on Regret: Analysis Through Hitchhiker, Car Accident, and Robbery Scenarios"
subtitle: "Reproduction of 'The impact of past behaviour normality on regret: replication and extension of three experiments of the exceptionality effect' (Lucas Kutscher & Gilad Feldman, 2018)"
author: Dailin(Ben) li, Qisheng Liang, Xutao Chen
thanks: "Code and data are available at: https://github.com/UofT-DailinLi/Impact-of-Atypical-Events-on-Regret.git. A replication of various aspects in this paper are available at: https://doi.org/10.1080/02699931.2018.1504747"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This study aims to replicate Kutscher and Feldman’s findings, which examine regret based on the norm theory. According to norm theory, an individual’s activity can be categorized as either typical or atypical， with the latter, on the other hand, eliciting a stronger emotional response. We focused on replicating three specific scenarios: hitchhiking, car accidents, and robberies. Examining how people recalled their behaviour in these specific scenarios helped our perception of the emotion of regret. This replication is consistent with the original study."

@@

toc: true
toc_depth: 2
number-sections: true
bibliography: references.bib
---

\newpage

```{r echo=FALSE, message=FALSE}

@@

<!-- Background / Broader Context -->

Regret is an uncomfortable state of mind that encompasses feelings of sorrow and distress over unfortunate events, perceived personal failures, losses, violations, deficiencies, or errors[@gilovich1995experience]. It associates to many things in our daily life. Matters of regret can include both acts done and opportunities missed; they may extend from intentional actions to those that are involuntary or happenstance; they could be tangible actions carried out or purely thoughts, whether self-initiated or by others; and they might constitute ethical or legal violations, or be actions that are neutral in terms of morality and law [@gilovich1995experience].

@@

-   Hitchhiker-scenario
-   Car accident scenario
-   Robbery scenario

@@

All of the data for REGRET and COMPENSATION contained 344 observations, and 21 VARIABLES were taken from the data underlying Kutscher and Feldman's study[@kutscher2019impact], which was collected as a questionnaire. The questionnaire recipients read a scenario in which a robbery occurred at a grocery store. In the first scenario (routine behaviour), Mr. Paul visited the store he frequented. In the second scenario (self-produced exception), the main character went to another store because he wanted a change of scenery. In the third scenario (other produced exception), Mr. Paul is forced to patronize another store because the store he frequents is closed for construction, and he runs into a robbery taking place in the store. In the next question, we ran a compensation measure followed by a regret measure. Respondents were asked how much money Mr. Paul should receive in compensation for his loss—scale from 0 to 10: 0\$ to 1000000\$. The next question was: assume no compensation was given to Mr. Paul. How much regret does he feel over the situation? Scale 1 from no regret to 5, very strong regret.

@@

## fig1

```{r echo=FALSE, message=FALSE, fig.cap="Proportions for perceived regret, injunctive social norms, descriptive social norms, and negative affect", label= "fig-1"}
fig1_data <- data.frame(
  Scenario = c("Smith - Routine", "Jones - Exception"),
  PerceivedRegret = c(7.9, 92.1),
  InjunctiveSocialNorms = c(4.7, 95.3),
  DescriptiveSocialNorms = c(90.6, 9.4),
  NegativeAffect = c(7.3, 92.7)
)

@@

p <- ggplot(data_long, aes(x = variable, y = value, fill = Scenario)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.7), width = 0.7) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, 100)) +
  labs(x = '', y = 'Percentage (%)', fill = 'Scenario') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = 'top')

@@

## The Influence of Exceptionality on Compensation and Regret

@@

Respondents were reading about whether Mr. Paul was injured because he happened to be robbed when he went to his everyday store or another store because of construction. Through the questionnaire, we observed the respondents on the compensation measure and the regret measure, and the two violin plots in @fig-3 show the distribution of the data. The distribution of compensation amount is roughly similar in regular and exceptional cases, and we can conclude that people do not define compensation separately for regular and special cases. However, we can observe that in the graph of the degree of regret, the degree of regret is more pronounced in the EXCEPTION situation. Regret is a complex emotional response; in general, regular events do not cause regret, but people are more likely to regret when they think that another decision might have led to a better outcome, which is in line with the results of our graph. Neal J. Roese and Amy Summerville have mentioned in their research that regret is related to perceived opportunities in life; that is, people perceive opportunities, the more opportunities people perceive, the more regretful they are[@roese2005we]. This is consistent with the situation in our study, in which Paul was robbed at a different store because of a store renovation, and respondents saw an opportunity not to be robbed, i.e., to go to the original store, and therefore had a generally higher level of regret. But when Paul is robbed when he goes to the original store, people don't see the chance not to be robbed, so the level of regret is smaller. So, our reproduced graph is consistent with the conclusions obtained from the original article.

# Conclusion

In summary, our study replicated Kutscher and Feldman's findings by examining the effects of past behaviour on regret, precisely scenarios related to hitchhiking, car accidents, and robberies. We found that individuals are more likely to experience regret when confronted with atypical events than with typical events. In hitchhiking and car accident scenarios, regret increases when people deviate from normal behavioural patterns. Also, we found that even the increase in regret does not affect the compensation domain. We can find from the violin plots that the need for compensation is very similar in both regular and special situations. Our analysis is consistent with the original research of Kutscher and Feldman and the findings of Neal J. Roese and Amy Summerville. It also emphasizes the complexity of regret as an emotion and that when people have more choices, they experience more regret when they believe they are missing out on other, more favourable outcomes because of the wrong choices[@roese2005we]. Future research could continue to explore the association between regret and other different emotions and whether levels of regret change over time, which could inform and help future psychology, compensation, and legal liability.
